Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Another pass, small fixes

\begin{abstract}
Union and intersection types are a staple of gradually typed language such
as TypeScript. While it's long been recognized that union and
intersections types are difficult to verify statically, it may appear
at first that the dynamic part of gradual typing is actually pretty
simple.

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding a static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\typescript{null} (\python{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, and its conformed of:

\begin{itemize}
  \item dictionaries, written as:
\begin{lstlisting}[language=nickel]
{field1 = value1, ..., fieldn = valuen}
\end{lstlisting}
  \item arrays:
\begin{lstlisting}[language=nickel]
[x1, x2, ..., xn]
\end{lstlisting}
  \item functions:
\begin{lstlisting}[language=nickel]
fun arg1 ... argn => body
\end{lstlisting}
  \item and let-definitions:
\begin{lstlisting}[language=nickel]
let id = value in exp
\end{lstlisting}  
\end{itemize}

@@

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were the original motivation for contracts).

@@

\subsection{Structure of this paper}
\unsure{Make this section into a contributions section?}
Section~\ref{sec:intro} introduced the perspective of designing a configuration
language. We presented the concept of contracts and detailed why contracts are
particularly fitted for a configuration language, and more generally for data
validation or program correctness in a dynamically typed language. The rest of
the paper is structured as follows:
\begin{itemize}
    \item Section~\ref{sec:feat-lang} outlines various common features of
        programming languages that may interact in a bad way with union and
        intersection contracts.
    \item Section~\ref{sec:union-inter} introduces unions and intersections and
        lists the possible motivations for adding them to a language.
    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
        encounter when trying to implement union and intersection contracts in
        practice.
    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
        existing attempts to solve said difficulties.
    \item Section~\ref{sec:related-work} provides some related work.
\end{itemize}

@@

\info{I think it's better to use [h] on figures, we can discuss it}
\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it where any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

@@

Another application of intersections shows up when intersecting
functions: it can be used to encode overloading.
For instance, take a look at Figure~\ref{fig:intersection-overloading-ex}, where the function
\nickel{duplicate} works both as a function to duplicate
arrays, as well as a function to duplicate strings.
This is particularly useful when using unions, since it's a
good way to express that a function can deal with different
shapes of data.
For instance, in the same figure, \nickel{duplicate} is (correctly) called
on a value of type \nickel{List Str @| Str}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let duplicate
    | (List Str -> List Str)
    @& (Str -> Str) =
  fun x => x ++ x in
let text | List Str @| Str = ... in
duplicate text
\end{lstlisting}
\caption{Duplicating an array of Strings or a String}
\label{fig:intersection-overloading-ex}
\end{figure}

@@

\subsubsection{case->}

>>>>

\begin{abstract}
Union and intersection types are a staple of gradually typed language such
as TypeScript. While it's long been recognized that union and
intersection types are difficult to verify statically, it may appear
at first that the dynamic part of gradual typing is actually pretty
simple.

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\typescript{null} (\python{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, augmented with abstraction mechanisms, and its conformed of:

\begin{itemize}
  \item dictionaries, written as:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
{field1 = value1, ..., fieldn = valuen}
\end{lstlisting}
  \item arrays:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
[x1, x2, ..., xn]
\end{lstlisting}
  \item functions:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
fun arg1 ... argn => body
\end{lstlisting}
  \item and let-definitions:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
let id = value in exp
\end{lstlisting}  
\end{itemize}

@@

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic checks
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were the original motivation for contracts).

@@

\subsection{Structure of this paper}
\unsure{Make this section into a contributions section?
Or maybe just kill it?}
Section~\ref{sec:intro} introduced the perspective of designing a configuration
language. We presented the concept of contracts and detailed why contracts are
particularly fitted for a configuration language, and more generally for data
validation or program correctness in a dynamically typed language. The rest of
the paper is structured as follows:
\begin{itemize}
    \item Section~\ref{sec:feat-lang} outlines various common features of
        programming languages that may interact in a bad way with union and
        intersection contracts.
    \item Section~\ref{sec:union-inter} introduces unions and intersections and
        lists the possible motivations for adding them to a language.
    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
        encounter when trying to implement union and intersection contracts in
        practice.
    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
        existing attempts to solve said difficulties.
    \item Section~\ref{sec:related-work} provides some related work.
\end{itemize}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it were any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

@@

\unsure{Maybe it's better to give it in Typescript syntax?
left it commented just in case}
% \begin{figure}[h]
% \begin{lstlisting}[language=JavaScript]
% type Animal = {
%   species: string,
%   breed:   string,
%   name:    string}

% type Pet = {owner: string, name: string}

% let myDog : Animal & Pet = {
%   species : "Canis Lupus",
%   breed   : "Australian Cattle Dog",
%   owner   : "Anonymous Author",
%   name    : "Juno" }
% \end{lstlisting}
% \caption{An animal that is also a pet}
% \label{fig:intersection-record}
% \end{figure}

Another application of intersections shows up when intersecting
functions: it can be used to encode overloading.
For instance, take a look at Figure~\ref{fig:intersection-overloading-ex}, where the function
\nickel{duplicate} works both as a function to duplicate
arrays, as well as a function to duplicate strings.
This is particularly useful when using unions, since it's a
good way to express that a function can deal with different
shapes of data.
For instance, in the same figure, \nickel{duplicate} is (correctly) called
on a value of type \nickel{(List Str) @| Str}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let duplicate
    | (List Str -> List Str)
    @& (Str -> Str) =
  fun x => x ++ x in
let text | (List Str) @| Str = ... in
duplicate text
\end{lstlisting}
\caption{Duplicating an array of Strings or a String}
\label{fig:intersection-overloading-ex}
\end{figure}

@@

\paragraph{case->}
